\pkg{dataMaid}: your assistant for data documentation in \proglang{R}

### 2017-10-19 latex/article_vol3.tex
Finished first full draft of new article version.

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{clean()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \code{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages of this data report are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{Example output from running \code{makeDataReport()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
   var1 var2  var3       var4 var5       var6
1   red    1     a -0.6264538    1 Irrelevant
2   red    1     a  0.1836433    2 Irrelevant
3   red    1     a -0.8356286    3 Irrelevant
4   red    2     a  1.5952808    4 Irrelevant
5   red    2     a  0.3295078    5 Irrelevant
6   red    6     b -0.8204684    6 Irrelevant
7   red    6     b  0.4874291    7 Irrelevant
8   red    6     b  0.7383247    8 Irrelevant
9   red  999     c  0.5757814    9 Irrelevant
10  red   NA     c -0.3053884   10 Irrelevant
11 blue    4     c  1.5117812   11 Irrelevant
12 blue   82     .  0.3898432   12 Irrelevant
13 blue   NA       -0.6212406   13 Irrelevant
14 <NA>  NaN other -2.2146999   14 Irrelevant
15 <NA>    5 OTHER  1.1249309   15 Irrelevant
\end{Soutput}
\end{Schunk}
\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData)
\end{Sinput}
\end{Schunk}

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the working directory and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data report consists of three
parts. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\smallskip Control input variables and summary\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isEmpty")} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{Inf} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, or pdf) & \code{"pdf"} \\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when the we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
10 printed potential errors for each variable. Also, we can add the
argument \code{replace=TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output = "html", onlyProblematic = TRUE, maxProbVals = 10,
+    replace = TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute proportion of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

\begin{Schunk}
\begin{Sinput}
R> allSummaryFunctions()
\end{Sinput}
\begin{Soutput}
----------------------------------------------------------------------------
name           description                     classes                      
-------------- ------------------------------- -----------------------------
centralValue   Compute median or mode          character, Date, factor,     
                                               integer, labelled, logical,  
                                               numeric

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
 	summaries = setSummaries(numeric = c("variableType","centralValue),
		integer = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
	summaries = setSummaries(all = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, checks = setChecks(factor = "identifyWhitespace"))
\end{Sinput}
\end{Schunk}

@@

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{checks} arguments, namely in the
pre-check stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()} and \code{isSingular()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, preChecks = "isKey")
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for each variable type. If, for instance, we wish to change the visualizations
from the default \pkg{ggplot2} \citep{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

@@

\section[Using dataMaid interactively]{\hl{Using \pkg{dataMaid} interactively}}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more
interactively through the data cleaning process. Aside from
the \code{clean()} function presented above, \pkg{dataMaid} also
provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or return the information
as an object for later use. This section describes how to use the
functions \code{check()}, \code{summarize()} and \code{visualize()} to
work interactively with \pkg{dataMaid}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \code{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\code{XXXChecks} argument (e.g., \code{numericChecks}, \code{factorChecks},
etc., as described in Section~\ref{sec:usingdataMaid}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \code{allCheckFunctions()} call provide
overviews of the available check functions.
Moving forward, we limit the numeric checks to only identify miscoded
missing values:

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the element
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{var2} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

\begin{Schunk}
\begin{Sinput}
R> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk}

@@

As we saw with the \code{check()} function, the summary can be modified
by providing the relevant \code{XXXSummaries} arguments. Setting the
\code{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\code{numericSummaries = c("centralValue", "minMax")}:

>>>>

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges, we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next up, we show how the data report and the interactive mode of \pkg{dataMaid} can be combined step-by-step to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we asses and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

Please note that this description of what \pkg{dataMaid} contains is usefully supplemented by a tutorial into how it can be extended. Every dataset is different, and some datasets might include problems that cannot be detected by our data checking functions. Therefore, \pkg{dataMaid} was designed to be easily extendend with user-supplied functions for summarizing, viusalizing and checking data. We have provided a vignette \hl{REF TO VIGNETTE} in which we describe how \pkg{dataMaid} extensions can be made, such that they are integrable with the \code{makeDataReport()}
function and with the other tools available in \pkg{dataMaid}.

@@

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf, html or word (.docx) format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages (excluding the frontpage) of this data report are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The two first pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. First a summary of the full dataset is given along with an overview of what checks were performed. Next, a summary of all the variables aand whether or not they are problematic is provided. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=4]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=5]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The third and fourth pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. Here, we see a description of each variable in the dataset, consiting of a summary table, a visualization and a indication of what problems were flagged for the variable (if any). At last, a few lines of metadata about the \code{makeDataReport()} are included for enhancing reproducibility. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example2}
\end{figure}

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
   var1 var2  var3       var4 var5       var6
1   red    1     a -0.6264538    1 Irrelevant
2   red    1     a  0.1836433    2 Irrelevant
3   red    1     a -0.8356286    3 Irrelevant
4   red    2     a  1.5952808    4 Irrelevant
5   red    2     a  0.3295078    5 Irrelevant
6   red    6     b -0.8204684    6 Irrelevant
7   red    6     b  0.4874291    7 Irrelevant
8   red    6     b  0.7383247    8 Irrelevant
9   red  999     c  0.5757814    9 Irrelevant
10  red   NA     c -0.3053884   10 Irrelevant
11 blue    4     c  1.5117812   11 Irrelevant
12 blue   82     .  0.3898432   12 Irrelevant
13 blue   NA       -0.6212406   13 Irrelevant
14 <NA>  NaN other -2.2146999   14 Irrelevant
15 <NA>    5 OTHER  1.1249309   15 Irrelevant
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData)
\end{Sinput}
\end{Schunk}

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
parts. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\smallskip Control input variables and meta information\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isSingular", "isSupported")} \\
\quad \code{reportTitle} & What should the title displayed on the front page of the report be? & \code{NULL} (corresponds to the dataset name) \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{"pdf"} \hl{OBS. Ændrer vi dette?}\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when the we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
2 printed potential errors for each variable. Also, we can add the
argument \code{replace=TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output = "html", onlyProblematic = TRUE,
+    maxProbVals = 2, replace = TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median for numeric variables, mode for categorical variables &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute proportion of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{makeDataReport()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

\begin{Schunk}
\begin{Sinput}
R> allSummaryFunctions()
\end{Sinput}
\begin{Soutput}
----------------------------------------------------------------------------
name           description                     classes                      
-------------- ------------------------------- -----------------------------
centralValue   Compute median for numeric      character, Date, factor,     
               variables, mode for             integer, labelled, logical,  
               categorical variables           numeric

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    summaries = setSummaries(numeric = c("variableType","centralValue),
+    integer = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    summaries = setSummaries(all = c("variableType", "centralValue"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    checks = setChecks(factor = "identifyWhitespace"))
\end{Sinput}
\end{Schunk}

@@

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{checks} arguments, namely in the
pre-check stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()}, \code{isSingular()} and \code{isSupported()}, check
whether a variable has unique values for all observations, only a
single value for all observations, and is not among the variable types supported by \pkg{dataMaid}, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow singular variables to move on to the SVC step by
only checking for keys and non-supported variables in the pre-check step:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, preChecks = c("isKey", "isSupported"))
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the report are also
controllable, though only a single function can be provided for each variable type. If, for instance, we wish to change the visualizations
from the default \pkg{ggplot2} \citep{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

@@

One might also encounter datasets with variables that are not among the 7 classes mentioned in the above (\code{character}, \code{Date}, \code{factor}, \code{integer}, \code{labelled}, \code{logical} and \code{numeric}), for instance variables of type \code{complex} or user-defined classes. It is possible to tell \code{makeDataReport()} how to handle such variables by use of the argument \code{treatXasY}. This argument takes a list where the names correspond to "new" variable types, while the entries must be supported variable types. For instance, we can instruct \code{dataMaid} to treat complex variables as numerics and generate a data report for a type \code{complex} variable like this:

\begin{Schunk}
\begin{Sinput}
>R complexData <- data.frame(complexVar = complex(100, real = 1:100, 
+    imaginary = 3), numericVar = 1:100))
>R makeDataReport(complexData, treatXasY(complex = "numeric"))
\end{Sinput}
\end{Schunk}

\section[Using dataMaid interactively]{Using \pkg{dataMaid} interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the data at various stages 
of the data cleaning process, it may be useful to be able to work more
interactively when performing actual data cleaning. Aside from
the \code{makeDataReport()} function presented above, \pkg{dataMaid} also
provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or return the information
as an object for later use. This section describes how to use the
functions \code{check()}, \code{summarize()} and \code{visualize()} to
work interactively with \pkg{dataMaid}.

@@

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \code{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the \code{checks} argument and the \code{setChecks()} helper function to specify this.
Recall that Table~\ref{table.SVCfunctions} or an \code{allCheckFunctions()} call provide
overviews of the available check functions.
Moving forward, we limit the numeric checks to only identify miscoded
missing values:

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2, checks = setChecks(numeric = "identifyMissing"))
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk} %$

@@

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the entry
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{var2} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

\begin{Schunk}
\begin{Sinput}
R> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk} %$

@@

As we saw with the \code{check()} function, the summary can be modified
by using the \code{summaries} argument and the \code{setSummaries()} helper function. If we want to remove the default summaries \code{variableType} and \code{countMissing} for numeric variables, we can use the function \code{defaultNumericSummaries()} and its argument \code{remove} that excludes a vector of summaries from the usual default summaries:
